---
title: "DO1200"
author: "Brian Yandell"
date: "2025-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rd_dir <- "/Volumes/mkeller3/General/main_directory/mapping_data"
dir(rd_dir)
```

```{r}
(project_info <- data.frame(project = "DO1200",
                           taxa = "CCmouse",
                           directory = "qtl2shinyData"))
```

```{r}
taxa_path <- file.path(project_info$directory, project_info$taxa)
(project_path <- file.path(taxa_path, project_info$project))
```

```{r}
if(!dir.exists(project_path)) dir.create(project_path, recursive = TRUE)
```

## Copy Key data for Clinical Example

### Objects from Cross object

```{r}
if(!file.exists(file.path(project_path, "covar.rds"))) {
  cross <- readRDS(file.path(rd_dir, "cross_DO_diet_grcm39_v6.rds"))
  cross$pheno <- NULL
  cross$geno <- NULL
  # Pull out individual objects.
  for(x in names(cross)) {
    savepath <- file.path(project_path, paste0(x, ".rds"))
    print(savepath)
    if(!file.exists(savepath)) {
      saveRDS(cross[[x]], savepath)
    }
  }
}
```

**Kinship is `kinship_loco` rather than `kinship`.**

### Genes and SNP Variants

```{r}
sql_files <- paste0(c("mouse_genes", "cc_variants"), "_grcm39.sqlite")
if(!all(file.exists(file.path(taxa_path, sql_files)))) {
  for(i in sql_files) {
    message("copying ", i)
    system(paste0("cp ", file.path(taxa_path, i), " ", taxa_path))
  }
}
```

Need to construct `query_variants.rds` and `query_genes.rds`.
Look on qtl2shiny vignette
[qtl2shinyData.Rmd](https://github.com/byandell-sysgen/qtl2shiny/blob/master/vignettes/qtl2shinyData.Rmd).
** Need to modify code to distinguish from old `gmc38` versions.**

```{r}
if(file.exists(qname <- file.path(taxa_path, "query_genes_grcm39.rds"))) {
  warning(paste("gene query", qname,
                "already exists and will not be overwritten"))
} else {
  query_genes <- 
    qtl2::create_gene_query_func(
      file.path(taxa_path, "mouse_genes_grcm39.sqlite"))
  saveRDS(query_genes, qname)
}
```

```{r}
if(file.exists(qname <- file.path(taxa_path, "query_variants_grcm39.rds"))) {
  warning(paste("variant query", qname,
                "already exists and will not be overwritten"))
} else {
  query_variants <- 
    qtl2::create_variant_query_func(
      file.path(taxa_path, "cc_variants_grcm39.sqlite"))
  saveRDS(query_variants, qname)
}
```

### Genotype Probabilities

Separate FST collections for alleles, allele pairs and SNPs.

```
probs_fst/
```

```{r}
list.files(file.path(rd_dir, "probs_fst"), pattern = "*.rds")
```

```{r}
if(!dir.exists(file.path(project_path, "probs_fst"))) {
  system(paste0("cp -r ", file.path(rd_dir, "probs_fst"), " ", project_path))
}
```

```{r}
dim(readRDS(file.path(project_path, "genoprob", "alleleprobs_fstindex.rds")))
```

```{r}
dim(readRDS(file.path(project_path, "genoprob", "genoprobs_fstindex.rds")))
```

```{r}
dim(readRDS(file.path(project_path, "genoprob", "snpprobs_fstindex.rds")))
```

Use `create_probs_query_func` from package 'qtl2pattern' to query genotype probabilities.
Save it in RDS format for later use. Note that it requires a relative address to the project data provided by `project_path`.

```{r}
if(file.exists(qname <- file.path(project_path, "query_probs.rds"))) {
  warning(paste("probs query", qname,
                "already exists and will not be overwritten"))
} else {
  query_probs <- qtl2pattern::create_probs_query_func(
    project_path, probdir = "probs_fst")
  saveRDS(query_probs, qname)
}
```

### Datasets and Phenotypes

Covariates were already retrieved from the `cross` object.
Phenotypes depend on `dataset`.

```{r}
datasets <- dir(file.path(rd_dir, "..", "files_for_cross_object"),
                pattern = "pheno_.*")
dataset_names <- stringr::str_replace(
  stringr::str_remove(
    stringr::str_remove(
      stringr::str_remove(
        stringr::str_remove(
          stringr::str_remove(datasets, ".csv$"),
          "_v[0-9]+$"),
        "_(DO_diet_batch|DO_diet|combat)_(corrected|nonzero150_qqnorm|nonzero150_vsd)"),
      "^pheno_"),
    "(in_vivo_|_GTT_metatraits)"),
  "((13C|2H))_metabolite", "metabolites_\\1")
names(datasets) <- dataset_names
datasets
```

The following saves `datasets` as `pheno_xxx.rds` with `xxx` the
`dataset_names`.
This is OK for now, but may want to save as FST later.
Also need to modify routine `pheno_read` or create another routine
`read_dataset`

```{r}
read_dataset <- function(project_info, dataset_name) {
  project_path <- paste(rev(project_info), collapse = "/")
  readRDS(file.path(project_path, paste0("pheno_", dataset_name, ".rds")))
}
```


```{r}
if(!file.exists(file.path(project_path, "pheno_clinical.rds"))) {
  for(i in dataset_names) {
    savepath <- file.path(project_path, paste0("pheno_", i, ".rds"))
    print(savepath)
    pheno_path <- file.path(rd_dir, "..", "files_for_cross_object", datasets[i])
    saveRDS(read.csv(pheno_path), savepath)
  }
}
```

### Peaks by Dataset and Covariates by Phenotype

There are many peaks files.
Need to decide if plan is to combine peaks files or have a bunch of
different ones. If combined, need columns for dataset, intset and dietset.

```
DO1200/peaks.rds
```

```{r}
peaksets <- dir(file.path(rd_dir, "..", "annotated_peak_summaries"),
    pattern = "^.*_all_.*additive_peaks.csv")
peakset_names <- stringr::str_replace(
  stringr::str_remove(
    stringr::str_remove(peaksets, "^DO1200_"),
    "_all_mice_additive_peaks.csv$"),
  "splice_juncs", "psi")
names(peaksets) <- peakset_names
peaksets
```

```{r}
peak_path <- file.path(project_path, "peaks")
for(i in tools::file_path_sans_ext(dir(peak_path))) {
  print(i)
  saveRDS(read.csv(file.path(peak_path, paste0(i, ".csv"))),
          file.path(peak_path, paste0(i, ".rds")))
}
```

```{r}
dir(file.path(rd_dir, "..", "annotated_peak_summaries"),
    pattern = "^.*_clinical_traits_.*additive_peaks.csv")
```

```{r}
dir(file.path(rd_dir, "..", "annotated_peak_summaries"),
    pattern = "^.*_clinical_traits_.*_peaks.csv")
```

```{r}
# "clinical_traits", liver_genes"m "liver_isoforms", "liver_lipids", "liver_splice_juncs"
dataset <- "clinical_traits"
# "all", "HC", "HF"
dietset <- "all"
# "additive", "interactive"
intset <- "additive"
peakpath <- file.path(rd_dir, "..", "annotated_peak_summaries",
                      paste0("DO1200_", dataset, "_", dietset,
                             "_mice_", intset, "_peaks.csv"))
print(peakpath)
savepath <- file.path(project_path, "peaks.rds")
print(savepath)
saveRDS(read.csv(peakpath), savepath)
```

Hotspots object is created with `hotspot`.
** Need to walk through this to find name changes in new data.**

```{r, eval=FALSE}
hots <- qtl2shiny::hotspot(pmap_obj,
                           dplyr::rename(peaks, lod = "qtl_lod"))
saveRDS(hots, file.path(project_path, "hotspot.rds"))
```

### Analyses Table

```
DO1200/analyses.rds.  # see comments below
```

The peaks files (previous section) include model information on
`addcovar` and `intcovar`.
It would be useful to switch to this.
Most of the rest of this is dated, I think.

The following need to be updated.
Phenotypes are easy, as we draw from the appropriate dataset.
Covariates--can use the whole data frame `covar`.

```{r}
pheno_data <- qtl2shiny:::pheno_read(project_info, analyses_tbl)
phe_mx <- qtl2shiny:::pheno_read(project_info, analyses_df)
cov_df <- qtl2mediate::get_covar(covar, analyses_df)
```

The `qtl2mediate::scan1covar()` function uses:

- `analyses_df$model`
- `which_covar(analyses_df)`
- `covarset <- apply(analyses_df, 1, function(x) paste(1 * x, collapse = ""))`
- `scanfn(probs_obj, phe_mx, kinship, cov_df, analyses_df, wh, models, ...)`

`which_covar` and `scanfn` are local functions

```{r}
scan_obj <- qtl2mediate::scan1covar(phe_mx, cov_df, probs_obj$probs, K_chr,
                                    analyses_df, sex_type = sex_type)
snp_scan_obj <- qtl2mediate::scan1covar(phe_mx, cov_df, snpprobs_act, K_chr,
                                analyses_df, sex_type = sex_type)
```

### SQL for genes and variants

```
CCmouse/mouse_genes_grcm39.sqlite
CCmouse/cc_variants_grcm39.sqlite
```

```{r, eval=FALSE}
for(i in c("mouse_genes", "cc_variants"))
system(paste0("cp ",
              file.path(rd_dir, paste0(i, "_grcm39.sqlite")),
              " ",
              taxa_dir))
```

Need to construct `query_variants.rds` and `query_genes.rds`.
Look on qtl2shiny vignette
[qtl2shinyData.Rmd](https://github.com/byandell-sysgen/qtl2shiny/blob/master/vignettes/qtl2shinyData.Rmd).

```{r}
if(file.exists(qname <- file.path(taxa_dir, "query_genes.rds"))) {
  warning(paste("gene query", qname,
                "already exists and will not be overwritten"))
} else {
  query_genes <- 
    qtl2::create_gene_query_func(
      file.path(taxa_path, "mouse_genes_grcm39.sqlite"))
  saveRDS(query_genes, qname)
}
```

```{r}
if(file.exists(qname <- file.path(taxa_dir, "query_variants.rds"))) {
  warning(paste("variant query", qname,
                "already exists and will not be overwritten"))
} else {
  query_variants <- 
    qtl2::create_variant_query_func(
      file.path(taxa_path, "cc_variants_grcm39.sqlite"))
  saveRDS(query_variants, qname)
}
```

### K `loco` matrix

```
DO1200/kinship.rds from k_loco_DO_diet_grcm39.rds
```

```{r, eval=FALSE}
system(paste0("cp ",
              file.path(rd_dir, "k_loco_DO_diet_grcm39.rds"),
              " ",
              file.path(project_path, "kinship.rds")))
```

```{r}
sapply(readRDS(file.path(rd_dir, "k_loco_DO_diet_grcm39.rds")), dim)
```

```{r}
sapply(readRDS(file.path(project_path, "kinship.rds")), dim)
```

### Maps and Covariates

```
DO1200/pmap.rds
DO1200/gmap.rds
DO1200/covar.rds
```

```{r}
cross <- readRDS(file.path(rd_dir, "cross_DO_diet_grcm39_v6.rds"))
```

```{r}
names(cross)
```

```{r}
sapply(cross$gmap, length)
sapply(cross$pmap, length)
```

```{r}
sum_na <- function(x) sum(!is.na(x))

# Generation (or wave) of each mouse.
table(cross$cross_info)

# Gene information; includes mus-human comparisons.
is_human <- grep("hum", names(cross$gene_annos))
sapply(cross$gene_annos[,-is_human], sum_na)
sapply(cross$gene_annos[is_human], sum_na)

# Isoform transcript information; includes mus-human comparisons.
is_human <- grep("hum", names(cross$transcript_annos))
sapply(cross$transcript_annos[,-is_human], sum_na)
sapply(cross$transcript_annos[is_human], sum_na)

# Genotype of founder across all markers (1,2,3)
sapply(cross$founder_geno, dim)

# Covariates; no missing values.
sapply(cross$covar, sum_na)
```

```{r}
save_crosser <- function(x, # name
                         frompath = fd_dir,
                         savex = x,
                         topath = project_path) {
  if(!file.exists(savepath <- file.path(topath, paste0(savex, ".rds")))) {
    saveRDS(cross[[x]], savepath)
  }
}
```

```{r, eval=FALSE}
save_crosser("pmap")
save_crosser("gmap")
```

```{r, eval=FALSE}
# Likely need to adapt to other covars as well?
# Gen = DOgen in code (DOwave also present)
# Check other covariates
save_crosser("covar")
```

```
> names(covar) # AttieDOv2
 [1] "mouse"              "sex"                "sac_date"           "partial_inflation" 
 [5] "coat_color"         "oGTT_date"          "FAD_NAD_paired"     "FAD_NAD_filter_set"
 [9] "crumblers"          "birthdate"          "diet_days"          "num_islets"        
[13] "DOgen"              "DOwave"
```

```
> names(cross$covar) # DO1200
 [1] "Gen"                    "GenLit"                 "Sex"                   
 [4] "Diet"                   "liver_lipids_batch"     "chol_batch"            
 [7] "true_TG_batch"          "KB_batch"               "plasma_13C_metab_batch"
[10] "plasma_2H_metab_batch"   
```

### Hotspots

```
DO1200/hotspot.rds
```

Hotspots object is created with

```{r, eval=FALSE}
hots <- qtl2shiny::hotspot(pmap, peaks)
saveRDS(hots, file.path(project_path, "hotspot.rds"))
```

### Phenotypes

```
DO1200/pheno_data.rds
```

This will take the most work to redesign.

### RNASeq

Need to review this carefully

```{r}

```

## Genoprobs for alleles, allele pairs and SNPs

```
probs_fst/
```

```{r}
list.files(file.path(rd_dir, "probs_fst"), pattern = "*.rds")
```

```{r}
dim(readRDS(file.path(rd_dir, "probs_fst", "alleleprobs_fstindex.rds")))
```

```{r}
dim(readRDS(file.path(rd_dir, "probs_fst", "genoprobs_fstindex.rds")))
```

```{r}
dim(readRDS(file.path(rd_dir, "probs_fst", "snpprobs_fstindex.rds")))
```
