---
title: "Prepare qtl2 data for qtl2shiny"
author: "Brian S. Yandell"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output: html_document
params:
  root_directory: "."
  shiny_directory: "qtl2shinyData"
  project: "Recla"
  taxa: "CCmouse"
  dontrun: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This documents how to use the `Recla` project data from <https://github.com/rqtl/qtl2data> with tools organized in [R/qtl2shiny](https://github.com/byandell/qtl2shiny).
See
[DeveloperGuide.Rmd](https://github.com/byandell-sysgen/qtl2shiny/blob/master/vignettes/DeveloperGuide.Rmd)
for creation of files.

The philosophy is to look in great depth at a small genome region
using real-time analyses and visualizations.

Code below uses internal functions `qtl2shiny:::*()`
that are helpers that keep the shiny code itself as simple as possible.
These can be examined to understand more of the nitty gritty.

The `project_info` data frame contains control information about each project (see [R/qtl2shiny](https://github.com/byandell/qtl2shiny)).

```{r eval=FALSE, echo = FALSE}
# For use in interactive mode for testing.
if(!exists("params") & interactive()) {
  params <- list(
    root_directory = "../../qtl2shinyDemo",
    shiny_directory = "qtl2shinyData",
    project = "Recla",
    taxa = "CCmouse",
    dontrun = FALSE)
}
```

```{r}
(project_info <- data.frame(project = params$project,
                           taxa = params$taxa,
                           directory = params$shiny_directory))
```

The processed data resides in the following directory:

```{r}
dir(project_info$directory)
```

If `dontrun` is `r TRUE` then just show remaining code without running.

```{r}
if(params$dontrun) {
  warning("Showing rest of code without running")
  knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
}
```

## Data setup

This selects a project, phenotype and chromosome position for
demonstration analysis.
Steps are from the
[setupServer](https://github.com/byandell-sysgen/qtl2shiny/blob/master/R/setupServer.R).

```{r}
project <- project_info$project
chr_id <- "11"
peak_Mbp <- 97
window_Mbp <- 1
pheno_name <- "LD_distance_light"
```

## Dashboard Setup

The dashboard uses the
[dashServer](https://github.com/byandell-sysgen/qtl2shiny/blob/master/R/dashServer.R)
to set up key data objects that are used throughout the app.
These are of two kinds: organizing tables and dataset.

Organizing tables have metadata that indicate how to use datasets.
These are legacy formats that need to be updated.

```{r}
analyses_tbl <- qtl2shiny:::read_project(project_info, "analyses")
analyses_df <- dplyr::filter(analyses_tbl, .data$pheno %in% pheno_name)
allele_info <- qtl2shiny:::read_project(project_info, "allele_info")
# Following used for mediation.
pheno_data <- qtl2shiny:::pheno_read(project_info, analyses_tbl)
```

Datasets are versions of data.
These will mostly be directly useable **except** that some names
may have changed over the years.

```{r}
phe_mx <- qtl2shiny:::pheno_read(project_info, analyses_df)
covar <- qtl2shiny:::read_project(project_info, "covar")
cov_df <- qtl2mediate::get_covar(covar, analyses_df)
K_chr <- qtl2shiny:::read_project(project_info, "kinship")[chr_id]
# Following used for mediation.
peaks <- qtl2shiny:::read_project(project_info, "peaks")
pmap_obj <- qtl2shiny:::read_project(project_info, "pmap")
```

## Haplotype Server

The
[haploServer](https://github.com/byandell-sysgen/qtl2shiny/blob/master/R/haploServer.R)
organizes the analyis using that allele haplotypes and the SNPs.
It can be used with different sex types, but only additive `"A"`
is illustrated here.

```{r}
sex_type <- "A"
```

The 
[probsServer](https://github.com/byandell-sysgen/qtl2shiny/blob/master/R/probsServer.R)
has three separate servers, for the 8 alleles (`probsServer`),
the 3 levels of SNPs (`snpProbsServer`),
and the 36 allele pairs (`pairProbsServer`).
Each of these has an API function `query_*()` to access
probability array by chromosome and `MBp` window.

```{r}
start_val <- peak_Mbp - window_Mbp
end_val <- peak_Mbp + window_Mbp
# Define query_probs function
query_probs <- qtl2shiny:::read_query_rds(project_info, "query_probs.rds")
# Note probs object keeps map with it
probs_obj <- query_probs(chr_id, start_val, end_val)
```

### Small Region Genome Scan

The
[scanCoefServer](https://github.com/byandell-sysgen/qtl2shiny/blob/master/R/scanCoefServer.R)
performs the scans.
For convenience and simplified reuse of key scan and coefficient objects,
we wrap
[qtl2::scan1](https://kbroman.org/qtl2/assets/vignettes/user_guide.html)
within
[qtl2mediate::scan1covar](https://github.com/byandell-sysgen/qtl2mediate/blob/master/R/scan1covar.R).


```{r}
scan_obj <- qtl2mediate::scan1covar(phe_mx, cov_df, probs_obj$probs, K_chr,
                                    analyses_df, sex_type = sex_type)
```

```{r}
scan_window <- c(start_val, end_val)
qtl2shiny:::plot_scan(scan_obj, probs_obj$map, seq(ncol(scan_obj)), chr_id, 
                      scan_window, phe_mx)
```

```{r}
blups <- FALSE
eff_obj <- qtl2shiny:::scan1_effect(probs_obj$probs, phe_mx, K_chr, cov_df,
                                    sex_type, blups)
qtl2shiny:::plot_eff(pheno_name, eff_obj, probs_obj$map, scan_obj, scan_window,
                 addlod = TRUE, allele_info)
```

### SNP Association and SDP Plots

`snpSetupServer`

```{r}
query_snp_probs <- qtl2shiny:::read_query_rds(project_info, "query_probs.rds")
snpprobs_obj <- query_probs(chr_id, peak_Mbp - window_Mbp,
                            peak_Mbp + window_Mbp, allele = FALSE)
      
      # define the query_variants function
query_variants <- qtl2shiny:::read_query_rds(project_info, "query_variants.rds")
snpinfo <- query_variants(chr_id, peak_Mbp - window_Mbp,
                          peak_Mbp + window_Mbp)
snpprobs_obj <- qtl2mediate::get_snpprobs(chr_id, peak_Mbp, window_Mbp,
  pheno_name, probs_obj$probs, probs_obj$map, snpinfo)
```

```{r}
snp_action <- "basic"
snpprobs_act <- qtl2pattern::snpprob_collapse(snpprobs_obj$snpprobs, snp_action)
snp_scan_obj <- qtl2mediate::scan1covar(phe_mx, cov_df, snpprobs_act, K_chr,
                                analyses_df, sex_type = sex_type)
```

```{r}
minLOD <- max(3, round(max(unclass(snp_scan_obj)), 1) - 1.5)
drop_hilit <- max(unclass(snp_scan_obj)) - minLOD
top_snps_tbl <- qtl2pattern::top_snps_pattern(snp_scan_obj,
  snpprobs_obj$snpinfo, drop_hilit)
```

Show max LOD for unique consequences.

```{r}
knitr::kable(
  dplyr::ungroup(
    dplyr::filter(
      dplyr::group_by(
        dplyr::select(
          top_snps_tbl,
#         dplyr::mutate(top_snps_tbl, consequence = abbreviate(consequence, 45)),
          snp_id, pos, sdp, lod, consequence),
        consequence),
      lod == max(lod), !is.na(consequence))))
```

```{r}
patterns <- 
  dplyr::arrange(
    dplyr::mutate(
      dplyr::filter(
        summary(top_snps_tbl), 
        .data$max_lod >= 3), 
      contrast = snp_action), 
    dplyr::desc(.data$max_lod))
```

`snpSetupServer`: `snpGeneServer`: `snpPlotServer`

```{r}
qtl2shiny:::top_snp_asso(snp_scan_obj, snpprobs_obj$snpinfo,
  scan_window, snp_action, minLOD = minLOD)
```

`snpSetupServer`: `snpPatternServer`

```{r}
dropHilit <- max(0, max(unclass(snp_scan_obj)) - minLOD)
qtl2shiny:::top_pat_plot(pheno_name, snp_scan_obj, chr_id, snpprobs_obj$snpinfo, 
  scan_window, drop_hilit = dropHilit, facet = "pattern",
  snp_action = snp_action)
```


`mediateServer`

```{r}
pos_Mbp <- peak_Mbp
qtls <- 1 # 1 or 2
other <- FALSE
med_plot_type <- "pos_LR"
local <- FALSE
signif <- TRUE
```

Assume comediator rather than expression.

```{r}
comed_ls <- qtl2mediate::comediator_region(pheno_name, chr_id, scan_window,
  covar, analyses_tbl, peaks, qtls, pmap_obj, pheno_data)
med_ls <- qtl2mediate::comediator_type(comed_ls, peaks, pheno_name, other)
```

```{r}
phe1_mx <- if(pheno_name %in% colnames(phe_mx)) {
    phe_mx[, pheno_name, drop = FALSE]
  } else {
    NULL
  }
mediate_obj <- qtl2mediate::mediation_test_qtl2(
          target = phe1_mx,
          mediator = med_ls[[1]],
          annotation = med_ls[[2]],
          covar_tar = cov_df,
          covar_med = med_ls$covar,
          genoprobs = probs_obj$probs,
          map = probs_obj$map,
          chr = chr_id,
          pos = pos_Mbp,
          kinship = K_chr)
```

```{r}
ggplot2::autoplot(mediate_obj, med_plot_type,
            local_only = local, 
            significant = signif) +
            ggplot2::geom_point(size = 4)
```

`triadServer`

```{r}
sdps <- unique(dplyr::filter(patterns, .data$pheno == pheno_name)$sdp)
haplos <- allele_info$code
choices_pattern <- qtl2pattern::sdp_to_pattern(sdps, haplos)
pattern <- choices_pattern[1]
sdp <- sdps[match(pattern, choices_pattern, nomatch = 1)]
medID <- "symbol"
choices_triad <- levels(shiny::req(mediate_obj)$best$triad)
triad <- choices_triad[1] 
choices_med <- dplyr::filter(mediate_obj$best, .data$triad == triad)$id
med_name <- choices_med[1]
```

```{r}
scat_dat <- qtl2mediate::mediation_triad_qtl2(
      target = phe_mx,
      mediator = med_ls[[1]][, med_name, drop = FALSE],
      annotation = med_ls[[2]],
      covar_tar = cov_df,
      covar_med = med_ls$covar,
      genoprobs = probs_obj$probs,
      map = probs_obj$map,
      chr = chr_id,
      pos = pos_Mbp,
      kinship = K_chr[[1]],
      sdp = sdp)
```

```{r}
peak_mar <- qtl2::find_marker(probs_obj$map, chr_id, pos_Mbp)
scat_plot <- "by_mediator"
```

** For some reason, the text is not showing on the
`qtl2shiny` rendition anymore. **

```{r}
ggplot2::autoplot(scat_dat, type = scat_plot,
             dname = peak_mar,
             mname = med_name,
             tname = colnames(phe_mx),
             fitlines = "sdp",
             centerline = NULL)
```

```{r}
ggplot2::autoplot(scat_dat, type = scat_plot,
             dname = peak_mar,
             mname = med_name,
             tname = colnames(phe_mx),
             fitlines = "sdp-parallel",
             centerline = NULL)
```

```{r}
ggplot2::autoplot(scat_dat, type = scat_plot,
             dname = peak_mar,
             mname = med_name,
             tname = colnames(phe_mx),
             fitlines = "parallel",
             centerline = NULL)
```

```{r}
ggplot2::autoplot(scat_dat, type = scat_plot,
             dname = peak_mar,
             mname = med_name,
             tname = colnames(phe_mx),
             fitlines = "driver",
             centerline = NULL)
```
